Visualizing demographic evolution using geographically inconsistent census data

### 2019-08-14 paper/intro.tex
first pass

\revision{In this work, we consider neighbourhoods as \emph{formal
regions}~\citep{montello2003regions}, geographically continous areas with
similar data characteristics.} These advances have made neighbourhood constructs
more tractable to quantitative analysis. Yet new challenges have also emerged,
especially at the convergence of research on neighbourhood effects and
neighbourhood dynamics. Neighbourhood effects research assumes knowledge about
the nature and scope of "the neighbourhood" that presumably shapes individual
outcomes~\citep{Kwan2018,Shelton2019}. Concurrently, researchers note that
neighbourhoods are not necessarily fixed containers in which other processes
occur, but themselves dynamically
evolve~\citep{Delmelle2017,Reades2019,li2018new}. The result is to open up key
assumptions about neighbourhoods for theoretical and empirical examination: how
do we appropriately define and compare neighbourhoods at a given time?; how do
we appropriately define and compare the temporal trajectories of
neighbourhoods?; and can we do both at once, "fully
interactionally"~\citep{Abbott1997}: classify neighbourhoods now based on where
they came from and where they are going?

@@

The standard approach then is to \emph{geographically harmonise} data. This
involves interpolating existing measurements into a common set of
regions~\citep{Logan2014,Hallisey2017,Allen2018}. Recent computational tools
have somewhat simplified this process~\citep{rey2018spatio}, but it still
involves non-trivial questions: which geometry to use as target, how to
apportion the variables, or how to combine data from different sources. Further,
these question do not necessarily have optimal answers. Indeed, regardless of
how well this process is performed, it still introduces
errors~\citep{Logan2016}, even when additional data is
provided~\citep{eicher2001dasymetric}. Essentially, harmonisation generates
artificial data points that can potentially lead to inaccurate results, even
though they are seldom interpreted as such. Nevertheless, because there has been
no viable alternative, and the results often appear plausible, these concerns
are generally overlooked. The result is that the harmonisation approach is
virtually mandatory in the current literature: \emph{"(...) tract-by-tract
comparison is not possible unless data from 2000 is interpolated to 2010
boundaries (...)"}~\citep{Dmowska2017}, \emph{"(...) This limits cross-year
comparison since data are not representative of the same spatial units. (...)"}
~\citep{Allen2018}.

@@

We start by presenting an intuitive example of our representation in
Section~\ref{sec:introduction}, then we review the relevant literature on
longitudinal studies, data representation, clustering, and spatio-temporal
visualisation in Section~\ref{sec:related}. Our methodology is introduced in
detail in Section~\ref{sec:method}, along with the included interface.
Illustrative scenarios for Chicago, Toronto, and Los Angeles are presented in
Section~\ref{sec:study} and the feedback of five field experts are summarised in
Section~\ref{sec:expert}. Our prototype system is available at
\censure{\url{http://uoft.me/piccard}}, including more than forty regions in the
US and Canada. The source code is publicly available at
\censure{\url{https://github.com/fabioasdias/piccard}}.\footnote{The editors are
considering, at our request, an exception to the double-blind requirement to
allow access to the system. We provided them with the URLs of the system, code,
and documentation separately.}

@@

To ease the cognitive transition to a new paradigm, we start with an intuitive
example of how the method works, using a small portion of a fictitious urban
region illustrated on the left part of Figure~\ref{fig:intuition}. This example
includes three different times ($t_0,t_1,t_2$), with different aggregation areas
identified as letters from A to H. For $t_0$, the initial time, we have areas A
and B, with small houses and a park, respectively. The park remains stable (B,
E, and H), but the houses are partially replaced by larger buildings (C and F).

This example illustrates some of the challenges of harmonisation. None of these
aggregation areas is clearly suitable as an interpolation target. In fact,
adopting any of these areas as a target would require merging heterogeneous
regions and/or dividing homogeneous regions. For instance, by choosing the
regions of $t_1$, A would be split to match regions C and D, which appears to be
a rather reasonable approximation in this homogeneous artificial example (even
if it commits the fallacy of division). Region F would be similarly split to
match C, but F would be split and merged with G to match D, potentially leading
to statistical measurements that do not properly represent either region.

@@

%I'm adding this because Jeff Allen (Farber's guy) struggled with this.
Note that the edges of this network merely encode that two regions are related.
This is binary information, there is no apportionment, no areal measurements,
no population percentages associated with the edge. Indeed, our method also
connects regions of the same time that share borders, representing exactly that
they are neighbouring areas.

In the following, we argue that this network representation allows us to study
neighbourhood change in a way that does not require an prior interpolation.

>>>>

Yet new challenges have also emerged, especially at the convergence of research
on neighbourhood effects and neighbourhood dynamics. Neighbourhood effects
research assumes knowledge about the nature and scope of "the neighbourhood"
that presumably shapes individual outcomes~\citep{Kwan2018,Shelton2019}.
Concurrently, researchers note that neighbourhoods are not necessarily fixed
containers in which other processes occur, but themselves dynamically
evolve~\citep{Delmelle2017,Reades2019,li2018new}. \revision{In this work, we
consider neighbourhoods as \emph{formal regions}~\citep{montello2003regions},
geographically continous areas with similar data characteristics, which these
advances made more tractable to approach in a data-driven fashion.} The result
is to open up key assumptions about neighbourhoods for theoretical and empirical
examination: how do we appropriately define and compare neighbourhoods at a
given time?; how do we appropriately define and compare the temporal
trajectories of neighbourhoods?; and can we do both at once, "fully
interactionally"~\citep{Abbott1997}: classify neighbourhoods now based on where
they came from and where they are going?

@@

The standard approach then is to \emph{geographically harmonise} the data. This
involves interpolating existing measurements into a common set of
regions~\citep{Logan2014,Hallisey2017,Allen2018}. Recent computational tools
have somewhat simplified this process~\citep{rey2018spatio}, but it still
involves non-trivial questions: which geometry to use as target, how to
apportion the variables, or how to combine data from different sources. Further,
these question do not necessarily have optimal answers. Indeed, regardless of
how well this process is performed, it still introduces
errors~\citep{Logan2016}, even when additional data is
provided~\citep{eicher2001dasymetric}. Essentially, harmonisation generates
\emph{artificial data points} that can potentially lead to inaccurate results,
even though they are seldom interpreted as such. Nevertheless, because there has
been no viable alternative, and the results often appear plausible, these
concerns are generally overlooked. The result is that the harmonisation approach
is virtually mandatory in the current literature: \emph{"(...) tract-by-tract
comparison is not possible unless data from 2000 is interpolated to 2010
boundaries (...)"}~\citep{Dmowska2017}, \emph{"(...) This limits cross-year
comparison since data are not representative of the same spatial units. (...)"}
~\citep{Allen2018}.

@@

We start by presenting an intuitive example of our representation in
Section~\ref{sec:introduction}, then we review the relevant literature on
longitudinal studies, data representation, clustering and regionalisation, and
spatio-temporal visualisation in Section~\ref{sec:related}. Our methodology is
introduced in detail in Section~\ref{sec:method}, along with the included
interface. Illustrative scenarios for Chicago and Toronto are presented in
Section~\ref{sec:study} and the feedback of five field experts are summarised in
Section~\ref{sec:expert}. Our prototype system is available at
\censure{\url{http://uoft.me/piccard}}, including more than forty regions in the
US and Canada. The source code is publicly available at
\censure{\url{https://github.com/fabioasdias/piccard}}.\footnote{The editors are
considering, at our request, an exception to the double-blind requirement to
allow access to the system. We provided them with the URLs of the system, code,
and documentation separately.}

@@

To ease the cognitive transition to this paradigm, we start with an intuitive
example of how the method works, using a small portion of a fictitious urban
region illustrated on the left part of Figure~\ref{fig:intuition}. This example
includes three different times ($t_0,t_1,t_2$), with different aggregation areas
identified as letters from A to H. For $t_0$, the initial time, we have areas A
and B, with small houses and a park, respectively. The park remains stable (B,
E, and H), but the houses are partially replaced by larger buildings (C and F).

The aggregation areas of none of these years is clearly suitable as an
interpolation target. Adopting any of them would require merging heterogeneous
regions and/or dividing homogeneous regions. For instance, by choosing the
regions of $t_1$, A would be split to match regions C and D, which appears to be
a rather reasonable approximation in this homogeneous artificial example (even
if it commits the fallacy of division). Region F would be similarly split to
match C, but F would be split and merged with G to match D, potentially leading
to statistical measurements that do not properly represent either region.

@@

%I'm adding this because Jeff Allen (Farber's guy) struggled with this.
Note that the edges of this network merely encode that two regions are related.
This is binary information, there is no apportionment, no areal measurements, no
population percentages, or weights of any kind associated with the edge. Indeed,
our method also connects regions of the same time that share borders,
representing exactly that they are neighbouring areas.

In the following, we demonstrate that this network representation allows us to
study neighbourhood change in a way that does not require prior interpolation.

### 2019-08-14 paper/method.tex
first pass

\section{Visualising the demographic spatio-temporal evolution}
\label{sec:method}
Figure~\ref{fig:overview} presents an overview of the processing steps of the
proposed method, illustrating how the nodes of the network are used to represent
the regions. The following sections elaborate this figure and explain the main
features of the interface we built to visualise and explore the evolution of
neighbourhoods on the basis of our proposed method.

@@

\subsection{Geographic content clustering}
Having tied the regions together into a network, we can now partition it to
identify similar sets of regions. We start by adopting a distance function
between the nodes, measuring the difference between the data of the regions.
This value is then associated with the edges, leading to a weighted dynamic
network. Every node has a collection of histograms, each representing the
distribution of certain aspect in the population.

@@

Once the distances are associated to the edges, we use watershed
cuts~\citep{Cousty2009} to create an initial clustering, which is then refined
into a hierarchy using the Sorted Maximal Matching (SMM)~\citep{markus2017} with
median linkage. The initial watershed step is performed to create an initial
clustering and reduce the running time of the SMM. We introduced one new
parameter to this method: a maximum distance threshold for the merges, to avoid
the early merging of outliers. We refer the reader to the original
paper~\citep{markus2017} for more details, including a complete performance
evaluation using several metrics. \revision{We chose this algorithm because it
is simple and easily customisable, and while different algorithms will lead to
different results, our methodology should work with any hierarchical clustering
algorithm.}

Each resulting cluster is contiguous in the network. This means that two
similar, but non-contiguous, sets of areas will be classified into two different
clusters, which can be counter-intuitive. To overcome this issue, we
\emph{augment} the network with two new edges per node from a nearest neighbours
graph~\citep{scikit-learn} using only the distances between the histograms.
These edges connect nodes with similar content, if they are not already
connected, providing a path for the algorithm to group similar nodes.
\revision{The regions connected by those edges will be merged on the first
stages of the clustering, since they are very similar, leaving the remaining
steps of the hierarchy to be determined only by the geographical edges. We
explored with different numbers of augmentation edges, but the results were not
consistent, since the distribution of the edges is data dependent. Adding two
edges per node was the least number of edges that led to stable and consistent
results in the scenarios available in our prototype. Since the problem of
balancing the data space with the geographical space is relevant for
geographical data analysis, this methodology potentially warrants further
exploration, beyond the scope of this work.}

>>>>

\section{Visualising the demographic spatio-temporal evolution}
\label{sec:method}
Figure~\ref{fig:overview} presents an overview of the processing steps of the
proposed method, illustrating how the nodes of the network are used to represent
the regions. The following sections elaborate on this figure and explain the
main features of the interface we built to visualise and explore the evolution
of neighbourhoods on the basis of our proposed method.

@@

\subsection{Geographic content clustering}
Having tied the regions together into a network, we can now partition it to
identify similar sets of regions. \revision{In essence, we are performing
temporal regionalisation by applying a regionalisation method over this
spatio-temporal network.}

We start by adopting a distance function between the nodes, measuring the
difference between the data of the regions. This value is then associated with
the edges, leading to a weighted dynamic network. Every node has a collection of
histograms, each representing the distribution of certain aspect in the
population.

@@

Once the distances are associated to the edges, we use watershed
cuts~\citep{Cousty2009} to create an initial clustering, which is then refined
into a hierarchy using the Sorted Maximal Matching (SMM)~\citep{markus2017} with
median linkage. The initial watershed step is performed to create an initial
clustering and reduce the running time of the SMM. We introduced one new
parameter to this method: a maximum distance threshold for the merges, to avoid
the early merging of outliers. We refer the reader to the original
paper~\citep{markus2017} for more details, including a complete performance
evaluation using several metrics. \revision{We chose this algorithm because it
is fast, simple, and easily customisable, but our methodology should work with
any hierarchical network clustering algorithm.}

Each resulting cluster is contiguous in the network. This means that two
similar, but non-contiguous, sets of areas will be classified into two different
clusters, which can be counter-intuitive. To overcome this issue, we
\emph{augment} the network with two new edges per node from a nearest neighbours
graph~\citep{scikit-learn} using only the distances between the histograms.
These edges connect nodes with similar content, if they are not already
connected, providing a path for the algorithm to group similar nodes.
\revision{The regions connected by those edges will be merged on the first
stages of the clustering, since the nodes they connect are, by definition, as
similar as possible\footnote{This assertion relies on the euclidean distance and
its relationship with the space where k-nn operates.}, leaving the remaining
steps of the hierarchy to be determined only by the geographical edges. We
experimented with different numbers of augmentation edges, but the results were
not consistent, since the distribution of the edges is data dependent. Adding
two edges per node was the smallest number of edges that led to stable and
consistent results in the scenarios available in our prototype. Since the
problem of balancing the data space with the geographical space is relevant for
geographical data analysis, this idea potentially warrants further exploration,
beyond the scope of this work.}

### 2019-08-14 paper/related.tex
first pass

Since our problem encompasses several fields, we divide this section into
specific sub problems: \emph{longitudinal demographic studies}, describing the
traditional tabular approach to longitudinal studies; \emph{data
representation}, elaborating how evolving geographic data can be represented for
processing; \emph{data clustering}, briefly reviewing existing clustering
methods; and \emph{cluster characterisation}, articulating how clusters can be
visually summarised.

@@

Network based representation of geographic information is fairly well explored
in the literature, as a basis for topological methods for event
detection~\citep{Doraiswamy2014}, leveraging signal processing on
graphs~\citep{shuman2013emerging,sandryhaila2013discrete} to find patterns and
outliers~\citep{Valdivia2015,Dias2015,Alce2018}. Networks are well suited to
represent trajectories as
well~\citep{VonLandesberger2016,Huang2016,chen2015survey}, allowing the use of
graph visualisation methods~\citep{Vehlow2015,Beck2014}. Our proposed method
builds upon this literature. We leverage a network-based representation that
removes the requirement for consistency in the measurement regions. Each region
in time corresponds to a different node. Instead of a collection of time-series,
the data is represented as a dynamic network.

@@

\revision{In geography, this problem is known as
\emph{regionalisation}~\citep{montello2003regions}, a rather old problem that
has been throughly explored, leveraging different mathematical tools, including
discrete topology~\citep{brantingham1978topological} and discrete
geometry~\citep{assunccao2006efficient}. Indeed, network-based methods are among
the current state-of-the-art~\citep{Guo2008Redcap,maxp}. However,
\emph{temporal} regionalisation is significantly less explored, especially in a
demographic context, arguably due to the difficulties in dealing with
unharmonised longitudinal data.   Recent neighbourhood related applications rely
on k-means~\citep{jain2010data,Delmelle2016}, the Louvain method for community
detection~\citep{blondel2008Louvain,Thomas2012}, or, to a lesser extent, Self
Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}, but do not
consider longitudinal data, or harmonised data.

Since we adopted a network-based data representation and our objectives include
an interactive interface, we opted for an heuristic variation of the maximum
weighted matching algorithm called \emph{sorted maximal
matching}~\citep{markus2017}, because of its simplicity, customisability, and
fast computation times. This algorithm merges clusters based on the weights of
the edges between pairs of clusters, creating an hierarchy over the data,
allowing the user to change the number of clusters without reprocessing the
data. Changing the clustering algorithm would lead to different results, but any
hierarchical network clustering method can be used in our framework. Indeed, our
core contribution is a different interpretation of the data representation, not
a new clustering or regionalisation method.}

>>>>

Since our problem encompasses several fields, we divide this section into
specific sub problems: \emph{longitudinal demographic studies}, describing the
traditional tabular approach to longitudinal studies; \emph{data
representation}, elaborating how evolving geographic data can be represented for
processing; \emph{data clustering and regionalisation}, briefly reviewing
existing data clustering methods,\revision{ geographic constraints and
regionalisation}; and \emph{cluster characterisation}, articulating how clusters
can be visually summarised.

@@

Network based representation of geographic information is fairly well explored
in the literature, as a basis for topological methods for event
detection~\citep{Doraiswamy2014}, leveraging signal processing on
networks~\citep{shuman2013emerging,sandryhaila2013discrete} to find patterns and
outliers~\citep{Valdivia2015,Dias2015,Alce2018}. Networks are well suited to
represent trajectories as
well~\citep{VonLandesberger2016,Huang2016,chen2015survey}, allowing the use of
network visualisation methods~\citep{Vehlow2015,Beck2014}. Our proposed method
builds upon this literature. We leverage a network-based representation that
removes the rigidity in the measurement regions. Each region in time corresponds
to a different node. Instead of a collection of time-series, the data is
represented as a dynamic network.

@@

\revision{In geography, this problem is known as
\emph{regionalisation}~\citep{montello2003regions}, a rather old problem that
has been throughly explored, leveraging different mathematical tools, including
discrete topology~\citep{brantingham1978topological} and discrete
geometry~\citep{assunccao2006efficient}. Indeed, network-based methods are among
the current state-of-the-art~\citep{Guo2008Redcap,maxp}. However,
\emph{temporal} regionalisation is significantly less explored, especially in a
demographic context, arguably due to the difficulties in dealing with
unharmonised longitudinal data. Recent neighbourhood related applications rely
on k-means~\citep{jain2010data,Delmelle2016}, the Louvain method for community
detection~\citep{blondel2008Louvain,Thomas2012}, or, to a lesser extent, Self
Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}.

Since we adopted a network-based data representation and our objectives include
an interactive interface, we opted for an heuristic variation of the maximum
weighted matching algorithm called \emph{sorted maximal
matching}~\citep{markus2017}, because of its simplicity, customisability, and
fast computation times. This algorithm operates on weighted networks, where a
distance metric is between the nodes is associated with the edges. We adopted a
distance based on the data associated with each node. The algorithm merges
clusters based on these distances, creating an hierarchy instead of a fixed
number of clusters. This hierarchical result is rendered by the interface,
allowing the user to change the number of clusters without reprocessing.
Changing the clustering algorithm would lead to different results, but any
hierarchical network clustering method, and distance metric, can be used in our
framework. }

### 2019-08-14 paper/study.tex
first pass

In this section we present two illustrative scenarios, using decennial census
data from the United States~\citep{nhgis} and
Canada\footnote{\url{http://datacentre.chass.utoronto.ca/census/}}, tabulated by
CTs,  from 1970 to 2010. \revision{Accidentaly, the US data for 2010 is actually
not from the decennial census, but from the ACS 2006-2010. Further, the regions
selected do not correspond to any pre-defined regions (metros, cities, census
areas), but to arbitrary regions defined around a location of interest. We
selected a minimal set of aspects for each country, and they are not similar to
each other, out of convenience, since the variable matching was a manual
process. These caveats would likely compromise any serious attempt on a
comprehensive demographic study, but this is not the objective of this
experiment. We aim only to test and validate our methodology, comparing our
results to pre-existing literature insights. Indeed, our results were perfectly
aligned to several other studies, despite these methodological missteps, which
could arguably be an indicator of the robustness of our method.}

@@

These results are meant to demonstrate the utility of the interface for
understanding the evolutionary dynamics of urban neighbourhoods. They also show
the face validity of the results generated by our novel network-based approach.

@@

We consider a region that is approximately the administrative border of the
current city of Toronto, using all seven available aspects with equal weights.
While Chicago was fairly stable, Toronto is known to be a more dynamic and
diverse city, with significant and increasing immigrant
population~\citep{hulchanski2007three,Fong2011}, especially
Asian~\citep{Fong2003}. Toronto is also known for a stable and well defined
Jewish community~\citep{Harold2018, Fong2011}. Therefore, we expect a
combination of stable and dynamic regions on the results, with Place of Birth,
Home Language, and Religion identified as relevant aspects. The results are
summarised in Figure~\ref{fig:to}, considering eight clusters.

@@

Indeed, the growing Asian population is visible starting in the 1980s and
building thereafter, leading to the yellow and pink clusters. While both include
a high percentage of people born in Asia, the pink is more defined by religion,
with low percentage of university degrees, and contains the lowest percentage of
people in the highest income bracket for these clusters; the yellow is less
defined by religion, and has higher education and income, geographically
corresponding to the Markham region, know for its Chinese population. A similar
division also happens for the two Jewish clusters, where the light green cluster
has lower education and income levels than the brown cluster. The purple cluster
of high income is somewhat stable. Until 2011 the cluster included the Bridle Path
neighbourhood, known for its wealthy population. In 2011 it was
classified into the yellow cluster of Asian immigration, since about 35\% of the
population for this CT were then born in Asia. The income distribution did not
change, with 85\% of the population with an income of 90k CAD or more.

@@

The experts also mentioned the poor responsiveness of the method when changes in
the clustering parameters required server-side processing (B,D). Indeed, the
current implementation can take a few minutes to cluster regions with high
number of CTs, like Los Angeles or Brooklyn. Server-processing reduced the
amount of data transferred to client, but it might increase the response time
under load. We implemented a cache policy that greatly improved the performance,
but fully pre-processing the results is not practical due to size of the
parameter space.

>>>>

In this section we present two illustrative scenarios, usingq census data from
the United States~\citep{nhgis} and
Canada\footnote{\url{http://datacentre.chass.utoronto.ca/census/}}, tabulated by
CTs,  from 1970 to 2010.

@@

\revision{Accidentaly, the US data for 2010 is actually not from the decennial
census, but from the ACS 2006-2010. Further, the regions selected do not
correspond to any pre-defined regions (metros, cities, census areas), but to
arbitrary regions defined around a location of interest. We selected a minimal
set of aspects for each country, and they are not similar to each other, out of
convenience, since the variable matching was a manual process. These caveats
would likely compromise any serious attempt on a comprehensive demographic
study, but this is not the objective of this experiment. These results are meant
to demonstrate the utility of the interface for understanding the evolutionary
dynamics of urban neighbourhoods. They also show the face validity of the
results generated by our novel network-based approach. Indeed, our results were
perfectly aligned with several other studies, despite these methodological
missteps, which could arguably be an indicator of its robustness.}

@@

We consider a region that corresponds approximately to the administrative border
of the current city of Toronto, using all seven available aspects with equal
weights. While Chicago was fairly stable, Toronto is known to be a more dynamic
and diverse city, with significant and increasing immigrant
population~\citep{hulchanski2007three,Fong2011}, especially
Asian~\citep{Fong2003}. Toronto is also known for a stable and well defined
Jewish community~\citep{Harold2018, Fong2011}. Therefore, we expect a
combination of stable and dynamic regions on the results, with Place of Birth,
Home Language, and Religion identified as relevant aspects. The results are
summarised in Figure~\ref{fig:to}, considering eight clusters.

@@

Indeed, the growing Asian population is visible starting in the 1980s and
building thereafter, leading to the yellow and pink clusters. While both include
a high percentage of people born in Asia, the pink is more defined by religion,
with low percentage of university degrees, and contains the lowest percentage of
people in the highest income bracket for these clusters; the yellow is less
defined by religion, and has higher education and income, geographically
corresponding to the Markham region, know for its Chinese population. A similar
division also happens for the two Jewish clusters, where the light green cluster
has lower education and income levels than the brown cluster. The purple cluster
of high income is somewhat stable. Until 2011 the cluster included the Bridle
Path neighbourhood, known for its wealthy population. In 2011 it was classified
into the yellow cluster of Asian immigration, since about 40\% of the population
for this CT were then born in Asia. The income distribution did not change, with
85\% of the population with an income of 90k CAD or more.

@@

The experts also mentioned the poor responsiveness of the method when changes in
the clustering parameters required server-side processing (B,D). Indeed, the
current implementation can take a few minutes to cluster regions with high
number of CTs, like Los Angeles or Brooklyn. Server-processing reduced the
amount of data transferred to client, but it might increase the response time
under load. We implemented a caching policy that greatly improved the
performance, but fully pre-processing the results is not practical due to size
of the parameter space.
